\section{GIT Directy Structure}~\label{sec:GIT}
On our beautiful github we have a lot of different projects : here is a litte overview of them.
https://github.com/Copilot-Language

\begin{enumerate}
\item Copilot : https://github.com/Copilot-Language/Copilot   This is the basic directory. Many branches but only two really active. Master and Struct. This repository contains all unit tests, basic documentation, and so on. Those two branches should be merged into master soon.
\item atom\_for\_copilot : https://github.com/Copilot-Language/atom\_for\_copilot   The modified package ATOM for generating ACSL contracts with Copilot. Some headers modified so NOT BACKWARD COMPATIBLE. One branch, probably deprecated.
\item sbv-for-copilot : https://github.com/Copilot-Language/sbv-for-copilot This used to be a modified package for SBV. Deprecated, DO NOT USE. All the changes have been merged into the official SBV version 5.0.
\item examplesForACSL : https://github.com/Copilot-Language/examplesForACSL  Some examples and projects in it to test. Makefiles provided. One branch : master. When using it, do not forget at each example to initialize the sandbox by cabal sandbox init --sandbox ../../Copilot/.cabal-sandbox/ 
\subitem Becareful ! Some examples are not fitted to be used with frama-c, or other tools (splint). Always check before running it, or you comnputer can crash by swapping.
\item copilot-cbmc : https://github.com/Copilot-Language/copilot-cbmc  insert description here TODO. One branch : master
\item copilot-c99 : https://github.com/Copilot-Language/copilot-c99  DEPRECATED : atom backend for SBV. Two branches. An original one, and an other using the atom\_for\_copilot modified package.
\item copilot-discussion : https://github.com/Copilot-Language/copilot-discussion Some blabla TODO
\item copilot-librairies : https://github.com/Copilot-Language/copilot-libraries  If you do not want to reinvent the wheel. 

\end{enumerate}

