\section{Extended Example: The Boyer-Moore Majority-Vote Algorithm}
\label{subsec:boyer_moore}

In this section we demonstrate how to use Haskell as an advanced macro language
on top of Copilot by implementing an algorithm for solving the voting problem
in Copilot.

\begin{figure*}[!htb]
\begin{lstlisting}[language = Copilot, frame = none]
majorityPure :: Eq a => [a] -> a
majorityPure []     = error "majorityPure: empty list!"
majorityPure (x:xs) = majorityPure' xs x 1

majorityPure' []     can _   = can
majorityPure' (x:xs) can cnt =
  let
    can' = if cnt == 0 then x else can
    cnt' = if cnt == 0 || x == can then succ cnt else pred cnt
  in
    majorityPure' xs can' cnt'
\end{lstlisting}
\caption{The first pass of the majority vote algorithm in Haskell.}
\label{fig:majority_pure}
\end{figure*}

\begin{figure*}[!htb]
\begin{lstlisting}[language = Copilot, frame = none]
aMajorityPure :: Eq a => [a] -> a -> Bool
aMajorityPure xs can = aMajorityPure' 0 xs can > length xs `div` 2

aMajorityPure' cnt []     _   = cnt
aMajorityPure' cnt (x:xs) can =
  let
    cnt' = if x == can then cnt+1 else cnt
  in
    aMajorityPure' cnt' xs can
\end{lstlisting}
\caption{The second pass of the majority vote algorithm in Haskell.}
\label{fig:amajority_pure}
\end{figure*}

Reliability in mission critical software is often improved by replicating
the same computations on separate hardware and by doing a vote in the end
based on the output of each system. The majority vote problem consists of
determining if in a given list of votes there is a candidate that has more
than half of the votes, and if so, of finding this candidate.

The Boyer-Moore Majority Vote Algorithm \cite{MooreBoyer82,Hesselink2005} solves
the problem in linear time and constant memory. It does so in two passes: The
first pass chooses a candidate; and the second pass asserts that the
found candidate indeed holds a majority.

Without going into details of the algorithm, the first pass can be implemented
in Haskell as shown in Figure \ref{fig:majority_pure}. The second pass, which
simply checks that a candidate has more than half of the votes, is
straightforward to implement and is shown in Figure \ref{fig:amajority_pure}.
E.g. applying {\tt majorityPure} on the string {\tt AAACCBBCCCBCC} yields {\tt
  C}, which {\tt aMajorityPure} can confirm is in fact a majority.

\begin{figure*}[!htb]
\begin{lstlisting}[language = Copilot, frame = none]
majority :: (Eq a, Typed a) => [Stream a] -> Stream a
majority []     = error "majority: empty list!"
majority (x:xs) = majority' xs x 1

majority' []     can _   = can
majority' (x:xs) can cnt =
  local
    (if cnt == 0 then x else can) $
      \ can' ->
        local (if cnt == 0 || x == can then cnt+1 else cnt-1) $
          \ cnt' ->
            majority' xs can' cnt'
\end{lstlisting}
\caption{The first pass of the majority vote algorithm in Copilot.}
\label{fig:majority}
\end{figure*}

\begin{figure*}[!htb]
\begin{lstlisting}[language = Copilot, frame = none]
aMajority :: (Eq a, Typed a) => [Stream a] -> Stream a -> Stream Bool
aMajority xs can = aMajority' 0 xs can > (fromIntegral (length xs) `div` 2)

aMajority' cnt []     _   = cnt
aMajority' cnt (x:xs) can =
  local
    (if x == can then cnt+1 else cnt) $
      \ cnt' ->
        aMajority' cnt' xs can
\end{lstlisting}
\caption{The second pass of the majority vote algorithm in Copilot.}
\label{fig:amajority}
\end{figure*}
% $

When implementing the majority vote algorithm for Copilot, we can use reuse
almost all of the code from the Haskell implementation. However, as functions
in Copilot are macros that are expanded at compile time, care must
be taken in order to avoid an explosion in the code size. Hence, instead of
using Haskell's built-in \emph{let}-blocks, we use explicit sharing, as
described in Section~\ref{sec:explicit_sharing}. The Copilot implementations
of the first and the second pass are given in Figure \ref{fig:majority} and
Figure \ref{fig:amajority} respectively. Comparing the Haskell implementation
with the Copilot implementation, we see that the code is almost identical,
except for the type signatures and the explicit sharing annotations.



