\section{Do No Harm} \label{sec:noninter} The RV components must be
composed with the SUO so that they are executing in parallel with the
SUO.  Care must be taken that the RV system itself does not compromise
the correct functioning of the SUO.  For instance executing monitors
may impact timing and scheduling. Care must be taken that any
instrumentation of the SUO does not affect the functional
correctness. In large systems, there are likely to be many monitors
running; each monitor might trigger different steering procedures. A common
pattern when things go wrong in complex safety-critical systems is
that many alerts are sounded simultaneously often placing a burden on
a human operator to sort things out.  One can easily envision an
analogous situation where several monitors detect violations,
triggering their respective steering procedures. Hence it is necessary
to verify that these different steering procedures do not interact with
each other in ways that could compromise safety. In summary,
high-assurance RV must uphold the Hippocratic oath``to do no harm''.
Ideally, we would formulate an noninterference theorem and the RV
framework would produce a proof certificate that the composed system
satisfies the property.


\paragraph{Challenge:} \emph{Assured RV must safely compose with the
  SUO.}

\paragraph{Copilot Approach:}  
The Copilot research group has yet to develop a general theory
noninterference, but we have made a number of design decisions with
this in mind, for instance, the choice of monitoring system state
through sampling was a deliberate attempt to minimize interference
with the SUO. Running monitors in separate partitions on an ARINC 653
compliant RTOS as in Section~\ref{sec:ft} ensures that any fault
in the RV will not negatively affect the executing SUO. The RTOS
scheduler also provides guarantees that a missed deadline in the RV
does not affect the SUO.

 

%\begin{itemize}
%\item Safety is known not to be compositional. 
%\item Keeping monitors independent help,but not a cure all. 
%\item Can the steering interact with the SUO to create problems. 
%\item Suppose I have a UAS with two monitors. One says if the vehicle  violates a
%geofence another says it's gone way off the reservation so fly into
%%the ground. 
%\item Maybe another example?
%\item Basically we have to make sure that the composition of SUO and
%monitors+steering don't impact safety.  
%\end{itemize} 