\section{Predictive RV} \label{sec:predictive}

\noindent\fbox{%
    \parbox{\textwidth}{
Question: Can the monitor detect a violation of the specification and invoke the safety controller in time to preserve safe operation.  
    }
}

Complex  real-time cyber-physical systems  are a natural target for
RV.  Applying RV to  functions such as an  adaptive control system
that have to act in strict time windows posses  technical
challenges. The monitor needs to detect  that the adaptive
controller is about to loose stability in time to  switch to a safe
controller. In the case of aircraft flight management, the RV system
needs to detect that  two airplanes are about to lose separation in
time for the aircraft to take corrective action.   Assured predictive
monitors are needed, but much work remains to be done. A promising
approach to monitoring controllers against a loss of stability is
given in~\cite{}. Researchers using Copilot have shown how  timing
analysis and  hard real-time scheduling of monitors can ensure that
monitors are executed at a set time.  Assured predictive monitoring
remains a research challenge for the RV community. 




%\item In the case of control s/w, the RV must detect a possible
%violation of stability in time to react. Mention Stan and Taylor's
%work.
%\item A flight management s/w it's hard to say. Need to take action if the
%FMS goes off and doesn't generate plans in time - a real possibility
%for genetic algorithms and what not. 
%\item Scheduling monitors in a RTOS so we know they at least execute
%at a certain rate is how we have tried to address this with our
%investigations of 653 RTOS.
%\end{itemize}