\section{Traceability}~\label{sec:trace}

\noindent\fbox{%
    \parbox{\textwidth}{%
Challenge : Is there traceability from the reqirememts and system level analysis to the actual monitor code?
    }%
}


If an IA system is the gain certification under existing guidance
through use of assured RV, the RV framework should generate the
supporting documentation.  An identifying characteristic of any of any
RV framework is the generation of monitors from
specifications. Typically, the RV transforms a safety property
expressed in logical notation such as LTL into a state machine encoded
in a traditional programming language. For instance, MoP translates
specifications into Java monitors while Copilot translates them into C
monitors. Approval of aviation software under the guidance of DO-178C
requires a traceability from requirements to object code. Guidance for
producing certified software in other safety-critical domains also
insist on traceability.  Industry has well established means for
establishing traceability from source code to object code.  To
facilitate the acceptance of RV in these areas, RV frameworks should
facilitate traceability from the monitor to the specification.  


Copilot's monitor generation process can create many small C
files. The code generation module has recently been revised to
generate code that is more readable and traceable.  We require the
user to insert labels in their specifications. The translation process
creates C header files with documentation formatted to be processed by
the plan text graph description language processor  DOT~\cite{ZZZZ}
giving  visual representation of the code  that includes the labels
making it easy to determine the correspondence between the monitor and
specification. 


%\begin{itemize}
%\item Many conventional processes based  guidance
% documents require ``traceability'' from the requirements to the
%  executable. 
%\item In assured RV we should be able to trace from requirements to
%  monitor spec to monitor. 
%\item We have developed a feature for adding labels to Copilot specs
%  facilitating traceability.  
%\item  We leveraged the GNU dot tool to help visualize what is going
%  on. 

%\end{itemize}
